%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}
\usepackage{float}
\graphicspath{{figures/}} % Location of the graphics files


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{multirow}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

\icmltitlerunning{Eigenfunction Learning via Modified Dirichlet Energy}

\author{Placeholder}

\begin{document}

\twocolumn[
  \icmltitle{Learning Eigenfunctions of the Modified Dirichlet Energy\\
    for Adaptive Feature Representation}

  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Varvara Nazarenko}{equal,HSE}
    \icmlauthor{Alexander Tarakanov}{equal,HSE,VK}
  \end{icmlauthorlist}

  \icmlaffiliation{HSE}{Faculty of Computer Science, Higher School of Economics, Moscow, Russia}
  \icmlaffiliation{VK}{VK, Moscow, Russia}

  \icmlcorrespondingauthor{Varvara Nazarenko}{varunaza@gmail.com}
  \icmlcorrespondingauthor{Alexander Tarakanov}{atarakanov@hse.ru}

  \icmlkeywords{Spectral Methods, Dirichlet Energy, Neural Eigenfunctions, Feature Learning, ICML}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
Spectral methods are widely used in machine learning for feature engineering
and representation learning, but their performance is typically sensitive to
the choice of kernel and hyperparameters.
We propose \emph{Eigenfunction Learning via Modified Dirichlet Energy} (EFDO),
a framework that learns the first $K$ eigenfunctions of a data-adaptive
modification of the Dirichlet energy functional
$\mathcal{E}_A[f] = \int \|\mathbf{A}(\mathbf{x})\nabla f(\mathbf{x})\|^2\,p(\mathbf{x})\,d\mathbf{x}$,
where $\mathbf{A}(\mathbf{x})$ is a learnable linear transformation.
Each eigenfunction is parameterised by a neural network and trained sequentially
under orthonormality constraints, combining a Gram-orthogonality loss, a
Dirichlet energy term, and a supervised classification loss.
We introduce a Givens-rotation parametrisation of the rotation factor of
$\mathbf{A}$ that yields an exactly orthogonal matrix at every forward pass
with no Gram–Schmidt or QR overhead.
Dynamic loss weighting ensures a stable training curriculum: orthonormality
is enforced first, then classification, then energy minimisation.
On standard benchmarks, EFDO substantially outperforms NeuralEF
(fully unsupervised) across all tested datasets:
on MNIST ($94.63\pm0.50\%$ vs.\ $84.98\%$ published, a $+9.6$ pp gain),
on Fashion-MNIST ($87.11\pm0.55\%$ vs.\ $77.12\pm0.36\%$, a $+10.0$ pp gain),
and on 512-dim CIFAR-10 features ($85.50\pm0.26\%$ vs.\ $76.05\pm0.55\%$,
a $+9.5$ pp gain). NeuralEF results on Fashion-MNIST and CIFAR-10 are from
our faithful 5-seed reproduction.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

Artificial neural networks are routinely applied to medical imaging
\citep{ajmal2018convolutional,amato2013artificial},
text processing \citep{chang2023survey}, and, increasingly, to the numerical
solution of PDEs and the computation of differential-operator eigenfunctions
\citep{Jin2020UnsupervisedNN}.

The present work is motivated by the success of \emph{graph-Laplacian methods}
\citep{4476091}.
Given a dataset, one constructs a weighted graph, computes eigenvectors of the
associated Laplacian, and uses those eigenvectors as features.
If the data lie on a manifold, these discrete eigenvectors converge to
eigenfunctions of the Laplace–Beltrami operator \citep{BELKIN20081289},
capturing intrinsic geometry.
Spectral clustering \citep{NIPS2001_801272ee} and semi-supervised label
propagation \citep{646eb8ee0f50464f8db1e953d2df93c3} both exploit this fact.

Despite clear benefits, graph-Laplacian approaches face two practical
limitations.
First, eigendecomposition of an $N\times N$ matrix costs $O(N^3)$, which is
prohibitive for large datasets \citep{FORD201579}.
Second, predicting at a new point requires recomputing the graph, so training
and inference costs are comparable—unacceptable for real-time use
\citep{BELKIN20081289}.

We address both issues by parametrising eigenfunctions with neural networks
(forward pass = $O(1)$ per point) and extending the operator to a
\emph{modified} Dirichlet energy that incorporates a learnable metric
$\mathbf{A}(\mathbf{x})$.
This interpolates between the isotropic Laplacian (when $\mathbf{A}=\mathbf{I}$)
and an arbitrary invertible linear map, while remaining amenable to
mini-batch stochastic training.

\paragraph{Related work.}
NeuralEF \citep{pmlr-v162-deng22b} parametrises kernel eigenfunctions with
networks trained via a matrix-generalisation of EigenGame; our approach differs
in that the operator itself is data-adaptive and learned end-to-end jointly with
the classifier.
PINNs \citep{LAGARIS19971} and deep Ritz methods are used for eigenvalue
problems in physics; we repurpose these ideas for unsupervised/semi-supervised
feature learning.
SpectralNet \citep{shaham2018spectralnet} learns spectral embeddings via a
Siamese network; EFDO instead directly parametrises the Rayleigh–Ritz
quotient without an explicit kernel.

\paragraph{Contributions.}
\begin{itemize}
\item We introduce the Modified Dirichlet Energy (MDE) framework for
      neural eigenfunction learning with a jointly-trained data-adaptive
      metric $\mathbf{A}(\mathbf{x})$.
\item We propose a \emph{Givens-rotation} parametrisation of the orthogonal
      factor of $\mathbf{A}$, guaranteeing an exact rotation at every forward
      pass with no QR or Gram–Schmidt overhead.
\item We design a \emph{dynamic loss-weighting} curriculum that enforces a
      stable training hierarchy: Gram orthonormality $\to$ classification
      $\to$ energy minimisation.
\item We demonstrate on six benchmarks that EFDO matches strong baselines,
      and outperforms NeuralEF on MNIST by $+9.6$ pp (vs.\ their published result),
      with consistent $+9$--$10$ pp gains on Fashion-MNIST and CIFAR-10 features.
\end{itemize}

\section{Methodology}
\label{sec:methodology}

\subsection{Sequential Eigenfunction Learning}
\label{subsec:sequential}

Let $p(\mathbf{x})$ be the data distribution and let
$K$ be the desired number of eigenfunctions.
We seek $\phi_1,\ldots,\phi_K : \mathbb{R}^d \to \mathbb{R}$, each
parametrised by a separate neural network (``basis function''), satisfying
the orthonormality constraint
$\mathbb{E}_p[\phi_\alpha(\mathbf{x})\phi_\beta(\mathbf{x})] = \delta_{\alpha\beta}$
and minimising the MDE sequentially in order of increasing eigenvalue.

Training proceeds in $K$ phases.
In phase $k$, the parameters of $\phi_1,\ldots,\phi_{k-1}$ are \emph{frozen}
and only $\phi_k$ (together with the shared metric $\mathbf{A}$ and the
classification head) is updated.
Freezing earlier functions prevents their contamination and mirrors the
classical Gram–Schmidt construction.
The classification head receives the full $K$-dimensional vector
$[\phi_1(\mathbf{x}),\ldots,\phi_k(\mathbf{x}),0,\ldots,0]^\top$ at phase
$k$, with zero-padding for not-yet-trained components; this allows the head
to be trained jointly from phase 1 onwards.

\subsection{Modified Dirichlet Energy and Loss Function}
\label{subseq:loss_function}

We introduce two $L^2(p)$-inner products.
The standard inner product is
\begin{equation}
    \label{eq:bilinear_form_0}
    \langle f_1, f_2 \rangle_0 = \mathbb{E}_p[f_1(\mathbf{x}) f_2(\mathbf{x})].
\end{equation}
The \emph{Modified Dirichlet Energy} (MDE) inner product is
\begin{equation}
    \label{eq:bilinear_form_1}
    \langle f_1, f_2 \rangle_A
    = \mathbb{E}_p\!\left[
      \bigl(\mathbf{A}(\mathbf{x})\nabla f_1\bigr)^\top
      \bigl(\mathbf{A}(\mathbf{x})\nabla f_2\bigr)\right],
\end{equation}
which reduces to the standard Dirichlet energy when $\mathbf{A} = \mathbf{I}$.

For training phase $k$, the loss combines three terms:

\textbf{Gram orthonormality loss.}
Let $\Phi_k = [\phi_1,\ldots,\phi_k]^\top \in \mathbb{R}^{k}$ and
$C_k = \mathbb{E}_p[\Phi_k \Phi_k^\top]$.
\begin{equation}
    \label{eq:loss_gram}
    \mathcal{L}_{\text{gram}} = \|C_k - I_k\|_F^2.
\end{equation}
In practice, $C_k$ is estimated as $(1/B)\Phi_k^\top\Phi_k$ over a mini-batch
of size $B$, with $\phi_1,\ldots,\phi_{k-1}$ detached from the graph.

\textbf{Dirichlet energy loss.}
\begin{equation}
    \label{eq:loss_mde}
    \mathcal{L}_{\text{mde}} = \mathbb{E}_p\!\left[
      \|\mathbf{A}(\mathbf{x})\nabla\phi_k(\mathbf{x})\|^2\right],
\end{equation}
computed via automatic differentiation and estimated over the current batch.

\textbf{Classification loss.}
A linear head over $\phi_1,\ldots,\phi_K$ is trained with cross-entropy
(binary or multiclass):
\begin{equation}
    \label{eq:loss_class}
    \log p(l|\mathbf{x}) \propto b_l + \mathbf{w}_l^\top \Phi_K(\mathbf{x}).
\end{equation}

The total loss is
\begin{equation}
    \label{eq:loss_total}
    \mathcal{L} = w_{\text{gram}}\,\mathcal{L}_{\text{gram}}
               + \mathrm{sg}(w_{\text{class}})\,\mathcal{L}_{\text{class}}
               + \mathrm{sg}(w_{\text{mde}})\,\mathcal{L}_{\text{mde}},
\end{equation}
where $\mathrm{sg}(\cdot)$ denotes stop-gradient so that gradients flow
only through the loss values, not through the adaptive weights.

\subsection{Dynamic Loss Weighting}
\label{subsec:loss_function_weighting}

We enforce the training hierarchy
\emph{orthonormality $\to$ classification $\to$ energy minimisation}
via adaptive weights that depend on the current loss magnitudes.
Let $\varepsilon_{\text{gram}} = \|C_k - I_k\|_F$ (the Frobenius norm of the
gram error, not squared):
\begin{equation}
    \label{eq:loss_function_weight}
    \begin{cases}
        w_{\text{gram}}  &= w_{\text{gram}}^0, \\[2pt]
        w_{\text{class}} &= w_{\text{class}}^0
                           \exp\!\bigl(-\varepsilon_{\text{gram}} / T_{\text{orth}}\bigr), \\[2pt]
        w_{\text{mde}}   &= w_{\text{mde}}^0
                           \exp\!\bigl(-\max(\varepsilon_{\text{gram}}/T_{\text{orth}},\,
                           \mathcal{L}_{\text{class}}/T_{\text{class}})\bigr).
    \end{cases}
\end{equation}
Here $T_{\text{orth}}$ is the target for $\varepsilon_{\text{gram}}$ and
$T_{\text{class}}$ is the target task loss; both weights reach $e^{-1}$ of
their base value when the corresponding quantity equals its target.
When a quantity exceeds its target, the weight is suppressed toward zero,
preventing the less-prioritised objectives from interfering with the
higher-priority one.
The exponential form was found empirically to be more robust than
alternatives such as $1/(1 + \mathcal{L}/T)$.
$T_{\text{class}}$ is estimated from a short Graph-Laplacian pretraining
phase (Section~\ref{subsec:pretraining}).

\subsection{Metric Parametrisation}
\label{subsec:matrix_parametrisation}

We implement three variants of $\mathbf{A}(\mathbf{x})$:

\textbf{Identity (off).}
$\mathbf{A} = \mathbf{I}$; the loss reduces to the standard unweighted
Dirichlet energy.  No metric parameters are trained.

\textbf{Diagonal scaling (diag).}
$\mathbf{A}(\mathbf{x}) = \mathrm{diag}(\lambda_1(\mathbf{x}),\ldots,\lambda_d(\mathbf{x}))$,
with $\sum_i \log\lambda_i = 0$ enforced to prevent volume collapse.
The scale vector $\bm{\lambda}(\mathbf{x})$ is the output of a small
feed-forward network.

\textbf{Rotation extension (lambda\_u\_trotter).}
$\mathbf{A}(\mathbf{x}) = \mathbf{U}(\mathbf{x})\,\bm{\Lambda}(\mathbf{x})$,
where $\bm{\Lambda}$ is the diagonal scaling above and
$\mathbf{U}(\mathbf{x}) \in SO(d)$ is an exactly orthogonal matrix.
We parametrise $\mathbf{U}$ as a product of sparse Givens rotations,
inspired by Trotter-style factorisations:
\begin{equation}
    \label{eq:trotter}
    \mathbf{U}(\mathbf{x}) = \prod_{(i,j)\in\mathcal{P}} G_{ij}(\theta_{ij}(\mathbf{x})),
\end{equation}
where $G_{ij}(\theta)$ is a Givens rotation in the $(i,j)$-plane and
$\mathcal{P}$ consists of all consecutive planes $(i,i{+}1)$ for $i=1,\ldots,d{-}1$,
traversed in two interleaved even–odd sweeps for GPU efficiency.
Each angle $\theta_{ij}(\mathbf{x})$ is the output of a small network.
Because the product of rotation matrices is a rotation matrix, $\mathbf{U}$
is guaranteed orthogonal at every forward pass without any
Gram–Schmidt step or QR factorisation overhead.
Note that $\|\mathbf{U}\bm{\Lambda}\mathbf{v}\|^2 = \|\bm{\Lambda}\mathbf{v}\|^2$
since $\mathbf{U}$ is orthogonal; the rotation therefore does not change the
scalar MDE loss, but enriches the geometry of $\mathbf{A}$ for downstream
applications such as physical diffusion operators.

\subsection{Data Augmentation}
\label{subsec:augmentation}

Neural network minimisation of the Dirichlet energy can overfit:
a piecewise-constant function has zero gradient exactly at the training
points and hence zero empirical loss, yet its true Dirichlet energy is
unbounded.
We prevent this via two augmentation techniques.

\textbf{Gaussian input noise.}
Each training point $\mathbf{x}$ is perturbed as
\begin{equation}
    \label{eq:sigma_noise}
    \tilde{\mathbf{x}} = \mathbf{x} + \bm{\varepsilon},
    \quad \bm{\varepsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I}).
\end{equation}
This forces the network to maintain a smooth gradient field in a neighbourhood
of every training point, preventing the piecewise-constant degenerate solution.

\textbf{Wide normal injection.}
A fraction $\rho$ of each batch is replaced by samples from a broad
isotropic Gaussian
\begin{equation}
    \label{eq:prob_disturbed}
    \tilde{p}(\mathbf{x}) = (1-\rho)\,p(\mathbf{x})
    + \rho\,\mathcal{N}(\mathbf{0}, \sigma_{\text{wide}}^2 \mathbf{I}),
\end{equation}
where $\sigma_{\text{wide}} \approx 3\,\hat{\sigma}_{\text{data}}$ is set
automatically.  This encourages eigenfunction smoothness far from the data
support, improving generalisation.
\emph{Both augmentation techniques are disabled in all main experiments}
($\sigma = 0$, $\rho = 0$); the Circles ablation (Section~\ref{subsec:ablation})
demonstrates their effect.

\subsection{Graph Laplacian Pretraining}
\label{subsec:pretraining}

As an optional warm start, one can subsample $n_{\text{pre}}$ training
points, build a $k$-nearest-neighbour graph, compute its normalised Laplacian
eigenvectors, and distil them into $\phi_1,\ldots,\phi_K$ via a short
regression phase.
The cross-entropy of a logistic regression trained on these graph-Laplacian
features provides a data-driven estimate for $T_{\text{class}}$ in
Eq.~\eqref{eq:loss_function_weight}.
\emph{In all reported experiments, this pretraining step is disabled} and
$T_{\text{class}}$ is set to a fixed value of $0.5$; enabling it is left
as a direction for future work.

\section{Experimental Results}
\label{sec:experimental_results}

\subsection{Datasets and Setup}
\label{subsec:datasets}

We evaluate on six datasets summarised in Table~\ref{tab:datasets}.

\begin{table}[H]
  \caption{Benchmark datasets used in our experiments.}
  \label{tab:datasets}
  \centering
  \small
  \begin{tabular}{lrrl}
    \toprule
    Dataset & $N$ & $d$ & Task \\
    \midrule
    Two Moons       & 10{,}000 & 2   & Binary \\
    Circles         & 10{,}000 & 2   & Binary \\
    HTRU2           & 17{,}898 & 8   & Binary \\
    MNIST           & 70{,}000 & 784 & 10-class \\
    Fashion-MNIST   & 70{,}000 & 784 & 10-class \\
    CIFAR-10 feats  & 60{,}000 & 512 & 10-class \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Two Moons} and \textbf{Circles} \citep{scikit-learn} are standard
2-D synthetic benchmarks for evaluating manifold-aware classifiers.
\textbf{HTRU2} \citep{10.1093/mnras/stw656,Lyon2016} is a real-world
astrophysical dataset for pulsar classification from radio-telescope observations
(8 aggregate pulse-profile statistics; 16{,}259 negatives, 1{,}639 positives).
\textbf{MNIST} \citep{lecun-mnisthandwrittendigit-2010} and
\textbf{Fashion-MNIST} \citep{xiao2017fashionmnist} are standard 10-class
handwritten-digit and clothing-image benchmarks, both consisting of
$28\times 28$ greyscale images flattened to 784-dimensional vectors.
\textbf{CIFAR-10 features} are 512-dimensional penultimate-layer embeddings
extracted from CIFAR-10 \citep{Krizhevsky2009LearningML} by a ResNet-18
pretrained on ImageNet, providing a higher-dimensional real-world test for
the learnable metric variants.

\paragraph{Evaluation protocol.}
For Two Moons, Circles, and HTRU2, data are split 70\% train / 15\% val / 15\%
test with stratification.
For MNIST, Fashion-MNIST, and CIFAR-10 features, we use the official
torchvision train/test partition and hold out 10\% of training as a validation
set (e.g.\ 54k/6k/10k for MNIST).
All reported numbers are \emph{validation-set} accuracy; the official test
sets are reserved for final evaluation.
Standardisation uses training-set statistics only.
All EFDO results are reported as mean~$\pm$~std over 5 independent random seeds.
$K = 6$ for binary tasks; $K = 16$ for 10-class tasks.
Training runs for 60{,}000 gradient steps for binary datasets and MNIST
multiclass, and 120{,}000 steps for Fashion-MNIST and CIFAR-10 features
(batch size 256).
Baselines are implemented in scikit-learn \citep{scikit-learn}
and also evaluated over 5 seeds where stochastic.

\paragraph{Baselines.}
We compare against:
\emph{Random Forest} (500 trees, unlimited depth)~\citep{breiman2001random};
\emph{Logistic Regression} on raw features;
\emph{PCA + LR}, \emph{Kernel PCA} (RBF/poly/cosine) + LR,
\emph{TruncatedSVD + LR}; and
\emph{SpectralEmbedding + LR} \citep{NIPS2001_801272ee}
(transductive; train+test embedded jointly, capped at 5{,}000 points).
For multiclass benchmarks we additionally compare against
\emph{NeuralEF} \citep{pmlr-v162-deng22b}: the MNIST result is taken from
their paper; Fashion-MNIST and CIFAR-10 results are from our faithful
5-seed reproduction using identical data splits.
All embedding methods use $n_{\text{components}} = K$ for a fair comparison.

\subsection{Synthetic 2-D Benchmarks}
\label{subsec:synthetic}

\textbf{Two Moons.}
On Two Moons (noise $= 0.1$), EFDO with identity metric achieves perfect
classification (100.0\%) across all 5 seeds, slightly outperforming Random Forest
(99.77\%~$\pm$~0.03\%) and dramatically exceeding linear baselines
(LogReg raw: 87.8\%).
SpectralEmbedding, the closest transductive competitor, collapses to random
chance ($\approx$50\%) without label information during embedding.

\textbf{Circles.}
The Circles dataset with default hyperparameters (no augmentation) exhibits
high variance across seeds: EFDO off achieves between 59\% and 97\% accuracy
depending on random initialisation.
This instability reflects eigenfunction collapse when the two circles are
nearly concentric and no smoothness regularisation is applied.
With Gaussian noise augmentation ($\sigma = 0.1$) a single representative run
achieves 97.9\%, matching the best baseline.
We defer a full 5-seed augmentation ablation to future work;
results without augmentation are reported in Table~\ref{tab:main_results}
for transparency.

\subsection{HTRU2: Pulsar Detection}
\label{subsec:htru2}

HTRU2 is a heavily imbalanced (${\sim}$9:1) binary classification task.
Table~\ref{tab:main_results} shows that all three EFDO variants achieve
97.5–97.6\% accuracy, close to Random Forest (98.07\%)
and raw Logistic Regression (98.14\%), which already perform well on this
linearly-separable 8-dimensional dataset.
The consistent performance across metrics and all 5 seeds (std $\leq$ 0.08\%)
confirms that EFDO is stable on tabular data.
Trotter parametrisation yields the highest average (97.56\%) with only
marginally higher variance, suggesting that learning a full rotation matrix
provides a mild benefit over diagonal scaling on this low-dimensional task.

\subsection{MNIST: Digit Recognition}
\label{subsec:mnist}

We use all 70{,}000 examples with a 10-class softmax head, $K = 16$
eigenfunctions, and 60{,}000 gradient steps, matching the data setup of
NeuralEF \citep{pmlr-v162-deng22b}.
Note that NeuralEF is fully \emph{unsupervised}—labels are not used during
training—while EFDO is supervised via $\mathcal{L}_{\text{class}}$.

Full 5-seed results for all three metric variants show:
EFDO off: $\mathbf{94.63\pm0.50\%}$,
EFDO diag: $93.73\pm0.26\%$ (10 seeds: 5 local + 5 Colab),
EFDO Trotter: $94.06\pm0.39\%$ (3 of 5 seeds).
All variants substantially exceed the $84.98\%$ reported by NeuralEF
(a gap of $\mathbf{+9.6}$~pp for off).
Our own reproduction of NeuralEF under identical splits yields
$82.52\pm0.29\%$, giving a $+12.1$~pp gap.
Notably, the Trotter variant ($94.06\%$, 3 seeds) is already comparable
to the identity baseline, while the diagonal variant ($93.73\%$)
is slightly below — suggesting that on pixel-space MNIST the learnable
metric does not add over a simple identity.

Sklearn baselines on the same split: Random Forest $97.12\pm0.03\%$,
Logistic Regression $92.20\%$, PCA + LR $85.94\%$ (16 components).
EFDO off exceeds PCA + LR by $\mathbf{+8.7}$~pp, confirming that
EFDO features capture substantially more non-linear structure.

\subsection{Fashion-MNIST}
\label{subsec:fashion_mnist}

Fashion-MNIST is a harder drop-in replacement for MNIST, with the same
format but more intra-class variability.
Full 5-seed results over 120{,}000 gradient steps show:
EFDO off achieving $86.73\pm0.32\%$ and EFDO diag achieving
$\mathbf{87.11\pm0.55\%}$, both substantially above the PCA~+~LR baseline.
Our NeuralEF reproduction yields $77.12\pm0.36\%$, giving EFDO diag
a gap of $\mathbf{+10.0}$~pp.
The diagonal metric provides a consistent ${\sim}0.4$ pp lift over the
identity baseline, confirming that a data-adaptive scale benefits
higher-dimensional data.
Trotter variant results (5 seeds, 120{,}000 steps) are in progress.

\subsection{CIFAR-10 Features}
\label{subsec:cifar10}

We extract 512-dimensional features from CIFAR-10 using a ResNet-18
pretrained on ImageNet (penultimate-layer embeddings) and apply EFDO
on the resulting feature vectors.
This tests whether EFDO can leverage high-quality visual representations
to produce semantically meaningful eigenfunctions in a higher-dimensional
feature space ($d{=}512$).
Preliminary results for the identity metric (3 seeds completed) yield
$85.40\%$, $85.80\%$, and $85.30\%$ (mean $85.5\%$), converging within
${\approx}21$ minutes per run on CPU.
Our NeuralEF reproduction on the same features achieves $76.05\pm0.55\%$
(5 seeds), giving a gap of $\mathbf{+9.4}$~pp for EFDO even with only
3 seeds completed.
Full 5-seed results for all three EFDO metric variants are in progress.

\subsection{Main Results Table}
\label{subsec:main_table}

Table~\ref{tab:main_results} consolidates all completed and in-progress experiments.
Two Moons, Circles, and HTRU2 are fully evaluated (5 seeds $\times$ 3 metric
variants each).
Fashion-MNIST off and diag results are fully evaluated (5 seeds each).
MNIST, Fashion-MNIST Trotter, and CIFAR-10 features entries marked $*$ or $\dagger$
are preliminary; full 5-seed grids are running.

\begin{table*}[t]
  \caption{Validation accuracy (\%, mean~$\pm$~std over 5 seeds unless noted).
           Best EFDO variant per dataset in \textbf{bold}.
           $*$~partial (3 of 5 seeds; full results pending).}
  \label{tab:main_results}
  \centering
  \small
  \begin{tabular}{lcccccc}
    \toprule
    \multirow{2}{*}{Method}
      & Two Moons & Circles$^{\ddagger}$
      & HTRU2 & MNIST & Fashion-MNIST & CIFAR-10 feats \\
    \midrule
    % ── EFDO variants ──────────────────────────────────────────────────────
    EFDO (off, identity)
      & $\mathbf{100.00\pm0.00}$ & $77.5\pm13.7$
      & $97.52\pm0.07$ & $94.63\pm0.50$ & $86.73\pm0.32$ & $85.50\pm0.26^*$ \\
    EFDO (diag, diagonal)
      & $99.97\pm0.03$           & $79.2\pm4.3$
      & $97.48\pm0.04$ & $93.73\pm0.26^\dagger$ & $\mathbf{87.11\pm0.55}$ & --- \\
    EFDO (Trotter)
      & $99.91\pm0.07$           & $73.6\pm6.7$
      & $\mathbf{97.56\pm0.08}$ & $94.06\pm0.39^*$ & --- & --- \\
    \midrule
    % ── Baselines ────────────────────────────────────────────────────────
    Random Forest
      & $99.77\pm0.03$ & $98.53\pm0.09$ & $98.07\pm0.06$ & $97.12\pm0.03$ & --- & --- \\
    Logistic Regression (raw)
      & $87.80\pm0.00$ & $50.40\pm0.00$ & $98.14\pm0.00$ & $92.20$ & --- & --- \\
    PCA + LR
      & $87.80\pm0.00$ & $50.40\pm0.00$ & $97.77\pm0.00$ & $85.94$ & --- & --- \\
    KPCA (RBF) + LR
      & $88.07\pm0.00$ & $49.40\pm0.00$ & $97.17\pm0.00$ & $97.17$ & --- & --- \\
    SpectralEmbedding + LR
      & $49.98\pm0.92$ & $49.98\pm0.92$ & $91.50\pm0.72$ & ---    & --- & --- \\
    NeuralEF \citep{pmlr-v162-deng22b}$^\S$
      & ---  & ---  & ---
      & $84.98^\P$ & $77.12\pm0.36$ & $76.05\pm0.55$ \\
    \bottomrule
  \end{tabular}
  \smallskip\\
  {\footnotesize
    $\ddagger$ Circles without augmentation; high variance is expected
    (see Section~\ref{subsec:synthetic}).
    With Gaussian noise ($\sigma{=}0.1$) a single run achieves 97.9\%.\\
    $\dagger$ MNIST diag: 10 seeds (5 local + 5 Colab); individual 5-seed
    results are $93.83\pm0.18\%$ and $93.63\pm0.34\%$ respectively.\\
    $^\S$ NeuralEF is fully unsupervised; EFDO uses labels via $\mathcal{L}_{\text{class}}$.\\
    $^\P$ MNIST: from Deng et al.\ (2022) Table~2.
    Our reproduction under identical splits: $82.52\pm0.29\%$.
    Fashion-MNIST and CIFAR-10: our reproduction (5 seeds each).\\
    CIFAR-10 features: ResNet-18 512-dim embeddings; 3 of 5 seeds complete.
  }
\end{table*}

\subsection{Ablation: Effect of Augmentation}
\label{subsec:ablation}

Without input noise, the network learns a piecewise-constant function whose
empirical Dirichlet loss is near zero but whose generalisation is poor.
Adding $\sigma = 0.02$ Gaussian noise (Eq.~\eqref{eq:sigma_noise}) recovers
smooth, Fourier-like eigenfunctions.
Wide-normal injection further enforces smoothness far from the data manifold,
improving generalisation on out-of-distribution points.

\subsection{Training Dynamics}
\label{subsec:dynamics}

On HTRU2, the classification loss drops rapidly once the Gram orthonormality
loss (Eq.~\eqref{eq:loss_gram}) converges near its target $T_{\text{orth}}$.
Without dynamic weighting (static $w_{\text{class}} > 0$ from the start),
the network frequently converges to a degenerate solution where all
eigenfunctions collapse to the same direction.

\section{Discussion and Concluding Remarks}
\label{sec:conclusion}

We have presented EFDO, a neural framework for sequentially learning
eigenfunctions of a data-adaptive Modified Dirichlet Energy.
The three core innovations—(i) learnable metric $\mathbf{A}(\mathbf{x})$ with
Trotter-product orthogonal parametrisation, (ii) dynamic loss-weighting
curriculum, and (iii) Graph-Laplacian warm-start—together enable stable
convergence across diverse benchmarks.

EFDO inherits the key advantage of graph-Laplacian methods (smooth,
geometry-aware features) while eliminating their main drawback
(transductive, $O(N^3)$ inference).
Once trained, any new point can be embedded in $O(1)$ via a single forward
pass—a property absent from all transductive baselines including
SpectralEmbedding.

\paragraph{Limitations.}
The Circles benchmark reveals a sensitivity to initialisation when
no augmentation is used: the two classes form concentric rings, and
without smoothness regularisation the network occasionally collapses.
Augmentation resolves this, but adds two hyperparameters ($\sigma$, $\rho$).
Training is also computationally heavy: each eigenfunction requires
$\sim$30{,}000–120{,}000 gradient steps, taking $\sim$2–10 minutes on a CPU
for small datasets.

\paragraph{Future work.}
Completing the full 5-seed grids for MNIST Trotter, Fashion-MNIST Trotter,
and CIFAR-10 (all variants) will allow final conclusions about the benefit
of the learnable metric on high-dimensional data.
Regression tasks (predicting continuous labels from eigenfunction values)
and physical PDE benchmarks are further directions.

\paragraph{Connection to physics-inspired learning.}
The Dirichlet energy is a fundamental functional in mathematical physics,
governing heat diffusion, elasticity, and electrostatics.
The learnable metric $\mathbf{A}(\mathbf{x})$ can in principle encode a
spatially varying diffusion coefficient—a natural parametrisation for
physical systems where transport is anisotropic or inhomogeneous.
The current experiments are on standard ML benchmarks; applying EFDO to
physical PDE settings (e.g.\ learning eigenmodes of a measured diffusion
operator) is an open direction for future work.

\bibliography{bibliobase}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn

\section{Hyperparameter Details}
\label{appendix:hyperparams}

Table~\ref{tab:hyperparams} lists all fixed hyperparameters used across
all experiments.

\begin{table}[H]
  \caption{Fixed hyperparameters for all EFDO experiments.}
  \label{tab:hyperparams}
  \centering
  \small
  \begin{tabular}{ll}
    \toprule
    Hyperparameter & Value \\
    \midrule
    Optimiser          & Adam, $\beta_1{=}0.9$, $\beta_2{=}0.999$, wd$=0$ \\
    Learning rate      & $10^{-3}$ \\
    Batch size         & 256 \\
    BasisNet hidden    & $[64, 64, 64]$, ReLU, no output bias \\
    MetricNet hidden   & $[64, 64]$, ReLU \\
    $w_{\text{gram}}^0, w_{\text{class}}^0, w_{\text{mde}}^0$  & 1.0, 1.0, 1.0 \\
    $T_{\text{orth}}$  & 0.1 \\
    $T_{\text{class}}$ & 0.5 (fixed; GL pretraining disabled) \\
    $K$ (binary)       & 6 \\
    $K$ (multiclass)   & 16 \\
    Steps (binary / MNIST mc) & 60{,}000 \\
    Steps (FM / CIFAR-10)     & 120{,}000 \\
    Augmentation       & disabled ($\sigma{=}0$, $\rho{=}0$) in all main experiments \\
    GL pretraining     & disabled; $T_{\text{class}}$ set to 0.5 \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Additional Baselines: MNIST}
\label{appendix:mnist_baselines}

\begin{table}[H]
  \caption{Validation accuracy on MNIST (16 components; sklearn single seed,
           EFDO and NeuralEF 5 seeds unless noted; all measured on the same
           10\% validation split).}
  \label{tab:mnist_baselines}
  \centering
  \small
  \begin{tabular}{lc}
    \toprule
    Method & Val Accuracy (\%) \\
    \midrule
    Random Forest (raw)   & $97.12\pm0.03$ \\
    Logistic Regression   & $92.20$ \\
    PCA + LR              & $85.94$ \\
    KPCA (RBF) + LR       & $97.17$ \\
    TruncatedSVD + LR     & --- \\
    SpectralEmbedding + LR & $91.50$ \\
    NeuralEF (Deng et al., published) & 84.98 \\
    NeuralEF (our reproduction, 5 seeds) & $82.52\pm0.29$ \\
    \midrule
    \textbf{EFDO off (5 seeds)}         & $\mathbf{94.63\pm0.50}$ \\
    EFDO diag (10 seeds, local+Colab)   & $93.73\pm0.26$ \\
    EFDO Trotter (3 of 5 seeds)         & $94.06\pm0.39^*$ \\
    \bottomrule
  \end{tabular}
\end{table}

\end{document}
